---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---

## Loading and preprocessing the data

```{r loadLibraries, echo=TRUE, message=FALSE}
#Loading the Libraries
library (knitr)
library(data.table)
library(ggplot2)
library(dplyr)
```
```{r setoptions, echo=TRUE, results='hide'}
#Setting the global options for knitr
opts_chunk$set(echo = FALSE, results = 'hide', tidy = TRUE)
```

Reading the Data:
```{r readData1, echo=TRUE, results='hide'}
#Reading the data
activity <- read.csv(file = 'D:/DataScience/ReproducibleResearch/Proj 1/Repo/activity.csv', stringsAsFactors = FALSE, sep = ',')
#Cleaning the data
activity$date <- as.Date(activity$date)
activity.complete <- activity[!is.na(activity$steps),] 
```
Below is a sample of the activity data
```{r activity_head, results = 'hold'}
head(activity, n = 7)
```
## What is mean total number of steps taken per day?
```{r mean_day, echo=TRUE, results='hold'}
steps_per_day <- aggregate(formula = steps~date, data = activity.complete,FUN = sum)
steps_per_day.mean <- round (mean(steps_per_day$steps),0)
steps_per_day.mean_text <- as.character(steps_per_day.mean)
steps_per_day.median <- round(median(steps_per_day$steps),0)
steps_per_day.median_text <- as.character(steps_per_day.median)
```

* The mean number of steps per day is `r steps_per_day.mean_text`.
* The median number of steps per day is `r steps_per_day.median_text`.

```{r hist_day, echo=TRUE, results='hold', fig.height=6}
hist(steps_per_day$steps, breaks = 15,main = 'Histogram of total number of steps taken each day', xlab = 'Total Number of Steps')
```

## What is the average daily activity pattern?

```{r time_series, echo=TRUE, results='hold'}
steps_per_interval <- aggregate(formula = steps~interval, data = activity.complete,FUN = mean)
steps_per_interval.max <- steps_per_interval[steps_per_interval$steps == max (steps_per_interval$steps),]
max_interval <- steps_per_interval.max$interval
max_interval.steps <- as.character(round(steps_per_interval.max$steps,0))
plot(steps_per_interval$interval, steps_per_interval$steps, type = 'l', main = 'Time Series', xlab = '5 minute time interval', ylab = 'Average steps per day')
```

The 5-minute interval of `r max_interval` has the maximum average of `r max_interval.steps`  steps per day.


## Imputing missing values

```{r missing_values_count, echo=TRUE, results='hold'}
missing_value_count <- sum (is.na(activity$steps))
```

The total number of missing values in the dataset are `r missing_value_count`.

Filling the missing values with the mean number of steps taken in the interval.
```{r full_activity, echo=TRUE, results='hold'}
activity.full <- merge(x = activity, y = steps_per_interval, by = "interval", all.x = TRUE)
s <- c('steps.x', 'steps.y')
activity.full[is.na(activity.full$steps.x),s[1]] <- activity.full[is.na(activity.full$steps.x),s[2]]
activity.full <- activity.full[,c('interval', 'date', 'steps.x')]
colnames(activity.full)[3] <- 'steps'
```
below is the extract of the filled dataset. It has not missing values
```{r filled_dataset, echo=TRUE, results='hold'}
head(activity.full, n = 7)
sum(is.na(activity.full$steps))
```
Calculating Mean, Median & Histogram based on the filled dataset:
```{r mean_day_full, echo=TRUE, results='hold'}
steps_per_day_full <- aggregate(formula = steps~date, data = activity.full,FUN = sum)
steps_per_day_full.mean <- round (mean(steps_per_day_full$steps),0)
steps_per_day_full.mean_text <- as.character(steps_per_day_full.mean)
steps_per_day_full.median <- round(median(steps_per_day_full$steps),0)
steps_per_day_full.median_text <- as.character(steps_per_day_full.median)
```

* The mean number of steps per day is `r steps_per_day_full.mean_text`.
* The median number of steps per day is `r steps_per_day_full.median_text`.

```{r hist_day_full, echo=TRUE, results='hold', fig.height=6}
hist(steps_per_day_full$steps, breaks = 15,main = 'Histogram of total number of steps taken each day', xlab = 'Total Number of Steps')
```

After filling in the missing data based on averages by interval, the is an insignificant change in the median. There is a small impact of imputing the missing data on the estimates of the total daily number of steps which does not seem to distort any decision making.

## Are there differences in activity patterns between weekdays and weekends?

```{r factor_weekday, echo=TRUE, results='hold'}
activity.full$day <- as.factor(ifelse(weekdays(activity.full$date) %in% c('Monday','Tuesday', 'Wednesday', 'Thursday', 'Friday' ), "Weekday", "Weekend")) 
steps_per_interval_day  <- aggregate(x = activity.full$steps ,by = list(activity.full$interval, activity.full$day), FUN = mean)
colnames(steps_per_interval_day) <- c ('interval', 'day', 'steps')
# build the plot
dayplot <- ggplot(steps_per_interval_day,aes(x=interval,y=steps, size=3)) + ggtitle('Time Series Plot of Average Steps by Interval & Day(weekday/weekend)') + facet_grid(. ~ day) + geom_line(size = 1)
dayplot
```